Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-18 paper/paper.tex
Uniformly write 'proof by logical relation' (without the s) for now

\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and a propositional equality type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

Concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We
  define the relation inductively, prove its functionality,
  define semantic typing in terms of the relation, and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Our proof
  showcases the special treatment required to model many common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply typed language. This shows
  that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs in Coq, with 983
  lines of code for the consistency proof with $\beta$ rules and a moderate
  increase to 1596 lines of code for the normalization proof with $\beta\eta$
  rules. We discuss our choice of using Coq as our metatheory,
  including the use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}. Our proof development is available as
  supplementary material.
\item We compare our work to existing proofs by logical relations and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of prior work (Section~\ref{sec:relatedwork}) and explain
  how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

@@

Not only does this close connection aid readers who wish to adopt proof
assistants for mechanizing metatheory, this precision is also important for
conveying the proof technique itself. Unlike properties derivable through
purely syntactic means, proofs by logical relations make demands on the
strength of the metatheory in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying metatheory requires
substantial effort from the reader to understand whether it is definable in a
given ambient logic.

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relations for
a dependently typed language with a predicative universe hierarchy, large
elimination, a propositional equality type, and dependent eliminators.
From the fundamental theorem for the logical relation, we prove consistency
and canonicity, then show the extensibility of our approach by proving the
existence of $\beta$($\eta$)-normal forms and adding natural numbers and
cumulativity to the language. These extensions only require small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that rivals pen and paper.

>>>>

\begin{abstract}
Proof by logical relation is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and a propositional equality type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

Concretely, our paper makes the following contributions.
\begin{itemize}
\item In Section~\ref{sec:spec}, we introduce \lang{}, the dependent type
  theory of interest. A key design choice that impacts our proofs is the use
  of an untyped conversion rule, inspired by Pure Type
  Systems~\citep{barendregt1991introduction}, and specified through parallel
  reduction~\citep{takahashi-parallel-reduction,
    barendregt:lambda-calculi-with-types}.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for
  \lang{}, the property of interest, to motivate a logical relation. We
  define the relation inductively, prove its functionality,
  define semantic typing in terms of the relation, and prove the
  fundamental theorem, from which consistency follows as a corollary
  (Section~\ref{sec:logrelproof}). Our proof
  showcases the special treatment required to model many common
  features of dependent type theories, thus making our proof applicable to a
  broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and $\beta\eta$-normal
  forms (Section~\ref{sec:betaeta}) for well-typed open terms. The
  modifications made to our initial logical relation are small and closely
  mirror the necessary extensions for a simply typed language. This shows
  that once we have established the base techniques, we can port
  ideas from proofs about simpler languages to the dependently typed setting.
\item We mechanize all our proofs in Coq, with 983
  lines of code for the consistency proof with $\beta$ rules and a moderate
  increase to 1596 lines of code for the normalization proof with $\beta\eta$
  rules. We discuss our choice of using Coq as our metatheory,
  including the use of off-the-shelf
  semantic engineering infrastructure and automation tools, in
  Section~\ref{sec:logrelmech}. Our proof development is available as
  supplementary material.
\item We compare our work to existing proofs by logical relation and other
  proof techniques for proving consistency and normalization.  We provide an
  overview of prior work (Section~\ref{sec:relatedwork}) and explain
  how various design decisions affect the size of our
  proof and its extensibility to additional features
  (Section~\ref{sec:discuss}).
\end{itemize}

@@

Not only does this close connection aid readers who wish to adopt proof
assistants for mechanizing metatheory, this precision is also important for
conveying the proof technique itself. Unlike properties derivable through
purely syntactic means, proofs by logical relation make demands on the
strength of the metatheory in which they are expressed. An informal proof that
attempts to be agnostic or ambiguous about the underlying metatheory requires
substantial effort from the reader to understand whether it is definable in a
given ambient logic.

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relation for
a dependently typed language with a predicative universe hierarchy, large
elimination, a propositional equality type, and dependent eliminators.
From the fundamental theorem for the logical relation, we prove consistency
and canonicity, then show the extensibility of our approach by proving the
existence of $\beta$($\eta$)-normal forms and adding natural numbers and
cumulativity to the language. These extensions only require small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that rivals pen and paper.
